---
title: "R Notebook"
author: Octavio del Ser
date: 10/11/2020
output: html_notebook
---

## Question 1
Generate three sets of 20 observation with their respective mean shift (60 total).
50 variables.
```{r}
num_classes <- 3
variables <- 50
obs <- 20
set.seed(2394)
total_obs <- obs * num_classes
x <- matrix(rnorm(num_classes*variables*obs), ncol = variables)
mean_shift <- sample(-10:10, obs, replace = FALSE)
for (var in 1:variables) {
  for (c in 0:(num_classes - 1)) {
    l<-(c * obs):((c + 1) * obs)
    x[l, var] <- x[l, var]+ mean_shift[c+1]
  }
}

```
## Question 2)
Perform k-means
```{r}

pr.out <- prcomp(x)
biplot(pr.out, scale = 0  )
```


## Question 3)
k-means -> k=3
```{r}
km.out <- kmeans(x, 3, nstart = 20)
plot(x, col = (km.out$cluster + 1), main = paste0("K-Means Clustering Results
with K=", num_classes, ' and mean shifts: ', paste(mean_shift, collapse = ', ')), xlab = "", ylab = "", pch = 20, cex = 2)
km.out
km.out$tot.withinss

l<-c(rep(1,20),rep(2,20),rep(3,20))
out<-unlist(km.out["cluster"])
table(l,out)
```
We can observer that the classification was correct, k means identified the mean shift in the dataset.
within cluster sum of squares by cluster result of 84.3%

## Question 4)
k-means -> k=3
Three distinct clusters were found. This indicates that there are were three principle components required to represent the data we generated. This is inline with the truth as
we created the data with three distinct mean shifts.

```{r}
set.seed(22)
km.out<-kmeans(pr.out$x[,c(1,2)],centers=3)

plot(pr.out$x[,c(1,2)], col = (km.out$cluster + 1), main = paste0("K-Means Clustering Results
with K=", num_classes, ' and mean shifts: ', paste(mean_shift, collapse = ', ')), xlab = "", ylab = "", pch = 20, cex = 2)
km.out
km.out$tot.withinss

l<-c(rep(1,20),rep(2,20),rep(3,20))
out<-unlist(km.out["cluster"])
table(l,out)
```