---
title: "R Notebook"
author: Octavio del Ser
date: 10/11/2020
output: html_notebook
---

##Question 1
Probability that first observation from the first bootstrap sample is not the jth observation
from original sample of size n:
This is the equivalent of 1-P(O1==Oj)
hence 1-(1/n)

##Question 2
Same as question 1 because samples are generated with replacement.


##Question 3
For the j'th probability to not be in the sample, it must not be in the first, second ... nth observation
in the Boostrap sample.
hence we multiply the probabilities (1-1/n) (not the first, second, third...) by each other so it becomes (not the first and not the second and not the third... ).
The result is (1-1/n)^n  power being number of observations in the given boostrap sample (which is the same as the number of observations in the set of observations)
```{r}
p <- function(n) {
  return((1 - (1 / n))^n)
}
```

##Question 4
```{r}
p(5)
```
##Question 5
```{r}
p(100)
```
##Question 6
```{r}
p(10000)
```
##Question 7
p(n) converges to ~ 0.3679 as n tends to infinity.
```{r}
n <- 1:100000
y <- p(n)
plot(n, y, xlab = 'n: 1:100,000', ylab = 'p(n) | p=(1-(1/n))^n)')
```
##Question 8
Error is acceptable
```{r}
store <- rep(0, 10000)
for (i in 1:10000) {
  store[i] <- sum(sample(1:100, rep = TRUE) == 4) > 0
}
r <- mean(store)
e <- 1 - p(100)

paste0("result:", r)
paste0("expected: ", e)
```
